---
title: "Basics of ggplot2"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)
```


## Setup

Load the necessary libraries

```{r}
library(tibble)
library(ggplot2)
library(tidyr)
library(readr)
```

## About `ggplot2`

A plot in `ggplot2` consists of several layers:

1. data
    - strict format: tidy data!
    - one row per datapoint
    - all grouping must be included in the data

2. aesthetics (`aes`) ([original documentation](https://ggplot2.tidyverse.org/reference/index.html#section-aesthetics))
    - how to map the data to the graph?
    - which column is *x*, *y*, ...
    - which column provides grouping (based on the grouping, one can either connect points into lines, split a graph into several facets, or color differently each group)

3. `geom`s ([original documentation](https://ggplot2.tidyverse.org/reference/index.html#section-geoms))
    - how to visualize the data?
    - points (`geom_point`), lines (`geom_line`), bars (`geom_bar`), etc.
    - `geom`s are connected to `stat`s that conduct any necessary pre-processing of data (e.g., `geom_histogram` would first calculate the number of observations in each bin through `stat_bin`)

4. `scale`s ([original documentation](https://ggplot2.tidyverse.org/reference/index.html#section-scales))
    - any type of data representation on the plot
    - coordinates: `scale_x_continuous`, `scale_x_discrete`, `scale_x_date`, etc.
    - colors: `scale_color_manual`, `scale_colour_brewer`, `scale_fill_continuous`, etc.

5. `theme` ([original documentation](https://ggplot2.tidyverse.org/reference/index.html#section-themes))
    - visual aspects
    - size and types of fonts
    - positioning of axes, legends, etc.

Each layer is added to the previous one with a `+` sign. The result can be saved to an object and added upon later.

### Example

Let's read in some data.

```{r}
hospital_inpatients <- read_csv("data/PasientSomatSykeh2.csv")
hospital_inpatients
```

This is publicly available data from SSB (Norwegian Statistics Bureau) on number of in-patients in hospitals in Hordaland between 2012 and 2018.

### Exercise 

*Here's a simple exercise with an empty code chunk provided for entering the answer.*

Write the R code required to add two plus two:

```{r two-plus-two, exercise=TRUE}

```

### Exercise with Code

*Here's an exercise with some prepopulated code as well as `exercise.lines = 5` to provide a bit more initial room to work.*

Now write a function that adds any two numbers and then call it:

```{r add-function, exercise=TRUE, exercise.lines = 5}
add <- function() {
  
}
```

## Topic 2

### Exercise with Hint

*Here's an exercise where the chunk is pre-evaulated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```
